---
title: "Data Manipulation"
author: "Ashkan"
output: 
  html_document: 
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
**ggplot** is a powerful and versatile data visualization package in R that has revolutionized the way data analysts and statisticians create high-quality plots. Developed by Hadley Wickham, ggplot follows the principles of the Grammar of Graphics, making it easy to construct complex visualizations by breaking them down into customizable components. With ggplot, users can effortlessly create aesthetically appealing and informative graphs by layering data, aesthetics, and geometric shapes. The package provides a wide range of customizable options for creating scatter plots, bar charts, line graphs, histograms, and more. The code-based approach of ggplot allows for a consistent and intuitive syntax, enabling users to efficiently modify and refine their visualizations. Whether for data exploration, presentation, or publication, ggplot continues to be a go-to tool for data visualization enthusiasts due to its flexibility, elegance, and seamless integration with R's data manipulation capabilities.


## Loading packages
```{r, warning=FALSE, message=FALSE}
library(gapminder)
library(car)
library(forcats)
library(tidyverse)
```

## Basics
```{r}
head(BOD)
```

```{r}
ggplot(data = BOD,
       mapping = aes(x = Time,
                     y = demand))+
  geom_point(size = 5)

ggplot(data = BOD,
       mapping = aes(x = Time,
                     y = demand))+
  geom_point(size = 5) +
  geom_line(color = "red")

ggplot(BOD, aes(Time, demand))+
  geom_point(size = 3)+
  geom_line(colour = "Red")
```

```{r}
head(CO2)
glimpse(CO2)
```

```{r}
CO2 %>%
  # x = and y = can be deleted
  ggplot(aes(x = conc, y =  uptake,
             color = Treatment))+
  geom_point()

CO2 %>%
  ggplot(aes(x = conc, y =  uptake,
             color = Treatment))+
  geom_point(size = 3, alpha = 0.5)+
  geom_smooth()

CO2 %>%
  ggplot(aes(x = conc, y =  uptake,
             color = Treatment))+
  geom_point(size = 3, alpha = 0.5)+
  geom_smooth(method = lm, se = F)

CO2 %>%
  ggplot(aes(x = conc, y =  uptake,
             color = Treatment))+
  geom_point(size = 3, alpha = 0.5)+
  geom_smooth(method = lm, se = F)+
  facet_wrap(~Type)

CO2 %>%
  ggplot(aes(x = conc, y =  uptake,
             color = Treatment))+
  geom_point(size = 3, alpha = 0.5)+
  geom_smooth(method = lm, se = F)+
  facet_wrap(~Type)+
  labs(title = "Concentration of CO2")+
  theme_bw()
```

```{r}
CO2 %>% 
  ggplot(aes(Treatment, uptake))+
  geom_boxplot()

CO2 %>% 
  ggplot(aes(Treatment, uptake))+
  geom_boxplot()+
  geom_point()

# Add mapping just for the point
# mapping can be done in original ggplot or
# in geometry with different reult
CO2 %>% 
  ggplot(aes(Treatment, uptake))+
  geom_boxplot()+
  geom_point(aes(size = conc,
                 colour = Plant))

CO2 %>% 
  ggplot(aes(Treatment, uptake))+
  geom_boxplot()+
  geom_point(alpha = 0.5
             ,aes(size = conc,
                 colour = Plant))

CO2 %>% 
  ggplot(aes(Treatment, uptake))+
  geom_boxplot()+
  geom_point(alpha = 0.5
             ,aes(size = conc,
                 colour = Plant))+
  coord_flip()

CO2 %>% 
  ggplot(aes(Treatment, uptake))+
  geom_boxplot()+
  geom_point(alpha = 0.5
             ,aes(size = conc,
                 colour = Plant))+
  coord_flip()+
  facet_wrap(~Type)+
  labs(title = "Title of the graph")+
  theme_bw()
```

```{r}
head(mpg)
glimpse(mpg)
```

```{r}
mpg %>% 
  ggplot(aes(displ, cty))+
  geom_point()

mpg %>% 
  ggplot(aes(displ, cty))+
  geom_point(aes(colour = drv,
                 size = trans),
             alpha = 0.5)

mpg %>% 
  ggplot(aes(displ, cty))+
  geom_point(aes(colour = drv,
                 size = trans),
             alpha = 0.5)+
  geom_smooth()

mpg %>% 
  ggplot(aes(displ, cty))+
  geom_point(aes(colour = drv,
                 size = trans),
             alpha = 0.5)+
  geom_smooth(method = lm)+
  facet_wrap(~year, nrow = 1)

mpg %>% 
  filter(cty < 25) %>% 
  ggplot(aes(displ, cty))+
  geom_point(aes(colour = drv,
                 size = trans),
             alpha = 0.5)+
  geom_smooth(method = lm)+
  facet_wrap(~year, nrow = 1)+
  labs(x = "Engine Size",
       y = "MPG in the city",
       title = "Fuel Efficiency")+
  theme_bw()

```

## Scatter plots
Two numerical variables
```{r}
mpg %>% 
  filter(hwy < 25) %>% 
  ggplot(aes(x = displ,
             y = hwy,
             colour = drv))+
  geom_point()+
  geom_smooth(method = lm,
              se = F)+
  facet_wrap(~year, nrow = 1)+
  labs(x = "Engine Size",
       y = "MPG on the Highway",
       title = "Fuel Efficiency")+
  theme_minimal()

mpg %>% 
  filter(hwy < 25) %>% 
  ggplot(aes(x = displ,
             y = hwy))+
  geom_point(aes(colour = drv))+
  geom_smooth(method = lm,
              se = F)+
  facet_wrap(~year, nrow = 1)+
  labs(x = "Engine Size",
       y = "MPG on the Highway",
       title = "Fuel Efficiency")+
  theme_minimal()

mpg %>% 
  filter(hwy < 25) %>% 
  ggplot(aes(x = displ,
             y = hwy))+
  geom_point(aes(colour = drv))+
  geom_smooth()+
  facet_wrap(~year, nrow = 1)+
  labs(x = "Engine Size",
       y = "MPG on the Highway",
       title = "Fuel Efficiency")+
  theme_minimal()
  
```

```{r}
# for multiple numeric variable scater plot is used
msleep %>% 
  ggplot(mapping = aes(x = bodywt, y = brainwt))+
  geom_point(aes(colour = sleep_total,
                 size = awake))+
  geom_smooth(method = lm, se = F)+
  labs(x = "Body Weight",
       y = "Brain Weight",
       title = "Brain and Body Weight")+
  theme_minimal()

msleep %>% 
  filter(bodywt < 2) %>% 
  ggplot(mapping = aes(x = bodywt, y = brainwt))+
  geom_point(aes(colour = sleep_total,
                 size = awake))+
  geom_smooth(method = lm, se = F)+
  labs(x = "Body Weight",
       y = "Brain Weight",
       title = "Brain and Body Weight")+
  theme_minimal()
```

## Bar charts and Histograms
One numerical variable (density, box plot, violin plots can be used)
```{r}
head(msleep)
glimpse(msleep)
```

```{r}

# for categorical variable bar chart
msleep %>% 
  drop_na(vore) %>% 
  ggplot(aes(x = vore))+
  geom_bar(fill = "#97B3C6")+
  theme_bw()+
  labs(x = "Vore",
       y = NULL,
       title = "Number of obseravtion per order")

msleep %>% 
  drop_na(vore) %>% 
  ggplot(aes(x = vore))+
  geom_bar(fill = "#97B3C6")+
  coord_flip()+
  theme_bw()+
  labs(x = "Vore",
       y = NULL,
       title = "Number of obseravtion per order")

msleep %>% 
  drop_na(vore) %>% 
  ggplot(aes(x = fct_infreq(vore)))+
  geom_bar(fill = "#97B3C6")+
  theme_bw()+
  labs(x = "Vore",
       y = NULL,
       title = "Number of obseravtion per order")

msleep %>% 
  drop_na(vore) %>% 
  ggplot(aes(x = fct_infreq(vore)))+
  geom_bar(fill = "#97B3C6")+
  coord_flip()+
  theme_bw()+
  labs(x = "Vore",
       y = NULL,
       title = "Number of obseravtion per order")
```

```{r}
# for numerical variable histogram
msleep %>% 
  ggplot(aes(x = awake))+
  geom_histogram()+
  theme_bw()+
  labs(x = "Tota; Sleep",
       y = NULL,
       title = "Histogram of total sleep")


msleep %>% 
  ggplot(aes(x = awake))+
  geom_histogram(binwidth = 2, fill = "#97B3C6")+
  theme_bw()+
  labs(x = "Tota; Sleep",
       y = NULL,
       title = "Histogram of total sleep")

```

## line graph
For two numeric and one categorical variables
```{r}
head(Orange)
glimpse(Orange)
```

```{r}
Orange %>% 
  filter(Tree != "2") %>% 
  ggplot(aes(age, circumference))+
  geom_point()+
  geom_smooth()+
  facet_wrap(~Tree)+
  theme_bw()+
  labs(title = "Tree age and Circumference")

Orange %>% 
  filter(Tree != "2" &
           Tree != "1") %>% 
  ggplot(aes(age, circumference, colour = Tree))+
  geom_point(size = 4, alpha = 0.3)+
  geom_line(linewidth = 1)+
  theme_minimal()+
  labs(title = "Tree age and Circumference")

```

## boxplot
One numeric one categorical variable
```{r}
msleep %>% 
  drop_na(vore) %>% 
  ggplot(aes(vore, sleep_total))+
  geom_boxplot()+
  theme_bw()

msleep %>% 
  drop_na(vore) %>% 
  ggplot(aes(vore, sleep_total))+
  geom_boxplot()+
  coord_flip()+
  theme_bw()

msleep %>% 
  drop_na(vore) %>% 
  ggplot(aes(sleep_total))+
  geom_histogram()+
  facet_wrap(~vore)+
  theme_bw()

msleep %>% 
  drop_na(vore) %>% 
  ggplot(aes(sleep_total))+
  geom_density()+
  facet_wrap(~vore)+
  theme_bw()

msleep %>% 
  drop_na(vore) %>% 
  ggplot(aes(sleep_total, colour = vore))+
  geom_density()+
  theme_bw()

msleep %>% 
  drop_na(vore) %>% 
  ggplot(aes(sleep_total, fill = vore))+
  geom_density(alpha = 0.2)+
  theme_bw()

msleep %>% 
  drop_na(vore) %>% 
  filter(vore %in% c("carni", "omni")) %>% 
  #filter(vore == "carni" | vore == "omni") %>% 
  ggplot(aes(sleep_total, fill = vore))+
  geom_density(alpha = 0.2)+
  theme_bw()
```

## Bar charts
Two categorical variables
```{r}
head(starwars)
glimpse(starwars)
```

```{r}
starwars %>% 
  filter(hair_color == "black" |
         hair_color == "brown") %>% 
  drop_na(sex) %>% 
  ggplot(aes(hair_color, fill = sex))+
  geom_bar(position = "dodge", alpha = 0.5)+
  theme_bw()

starwars %>% 
  filter(hair_color == "black" |
         hair_color == "brown") %>% 
  drop_na(sex) %>% 
  ggplot(aes(hair_color, fill = sex))+
  geom_bar(position = "dodge", alpha = 0.5)+
  theme_bw()+
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())+
  labs(title = "Gender and Hair colour",
       x = "Hair Colur",
       y = "Number")

starwars %>% 
  filter(hair_color == "black" |
         hair_color == "brown") %>% 
  drop_na(sex) %>% 
  ggplot(aes(hair_color, fill = sex))+
  geom_bar(alpha = 0.5)+
  theme_bw()+
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())+
  labs(title = "Gender and Hair colour",
       x = "Hair Colur",
       y = "Number")

starwars %>% 
  filter(hair_color == "black" |
         hair_color == "brown") %>% 
  drop_na(sex) %>% 
  ggplot(aes(hair_color, fill = sex))+
  geom_bar(position = "fill", alpha = 0.5)+
  theme_bw()+
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())+
  labs(title = "Gender and Hair colour",
       x = "Hair Colur",
       y = "Number")

starwars %>% 
  filter(hair_color %in% c("black", "brown")) %>% 
  drop_na(sex) %>% 
  ggplot(aes(sex))+
  geom_bar(aes(fill = sex), alpha = 0.5)+
  facet_wrap(~ hair_color)+
  theme_bw()+
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())+
  labs(title = "Gender and Hair colour",
       x = "Hair Colur",
       y = "Number")

```

## some example
```{r}
head(gapminder)
glimpse(gapminder)
```

```{r}
gapminder %>% 
  filter(continent %in% c("Africa", "Europe")) %>% 
  filter(gdpPercap < 30000) %>% 
  ggplot(aes(x = gdpPercap,
             y = lifeExp,
             size = pop,
             color = year))+
  geom_point()+
  facet_wrap(~continent)+
  labs(title = "Life expectancy explained by GDP",
       x = "GDP per capita",
       y = "Life expectancy")

```

##Error Bars
```{r}
head(ToothGrowth)
glimpse(ToothGrowth)
```

```{r}
ToothGrowth %>% 
  filter(supp == "VC") %>% 
  mutate(dose = as.factor(dose)) %>% 
  group_by(dose) %>% 
  summarise(mean_len = mean(len),
            sd_len = sd(len))
 
ToothGrowth %>% 
  filter(supp == "VC") %>% 
  mutate(dose = as.factor(dose)) %>% 
  group_by(dose) %>% 
  summarise(mean_len = mean(len),
            sd_len = sd(len)) %>% 
  ggplot(aes(x = dose, y = mean_len))+
  geom_point(size = 3,
             colour = "Blue")+
  geom_errorbar(aes(x = dose,
                    ymin = mean_len - sd_len,
                    ymax = mean_len + sd_len,
                    width = 0.05))+
  labs(title = "Average tooth growth for VC supplement group")+
  xlab("Dose")+
  ylab("Average tooth growth")+
  theme_bw()+
  theme(panel.grid.major = element_blank())

ToothGrowth %>% 
  filter(supp == "VC") %>% 
  mutate(dose = as.factor(dose)) %>% 
  group_by(dose) %>% 
  summarise(mean_len = mean(len),
            sd_len = sd(len)) %>% 
  ggplot(aes(x = dose, y = mean_len))+
  geom_bar(stat = "identity",
           fill = "skyblue",
           alpha = 0.5)+
  geom_errorbar(aes(x = dose,
                    ymin = mean_len - sd_len,
                    ymax = mean_len + sd_len,
                    width = 0.05))+
  labs(title = "Average tooth growth for VC supplement group")+
  xlab("Dose")+
  ylab("Average tooth growth")+
  theme_bw()+
  theme(panel.grid.major = element_blank())
```

## Lollipop Graph
Between a categorical and numerical variable
```{r}
msleep %>% 
  group_by(order) %>% 
  summarise(mean_sleep = mean(sleep_total)) %>% 
  mutate(order = fct_reorder(order, mean_sleep)) %>% 
  ggplot(aes(x = order, y = mean_sleep))+
  geom_point(size = 3,
             colour = "orange")+
  geom_segment(aes(x = order, # x of the starting point of all line
                   y = mean(msleep$sleep_total), # y of starting point
                   xend = order,  # x of the end points of lines
                   yend = mean_sleep),
               colour = "gray")+
  geom_hline(yintercept = mean(msleep$sleep_total),
             colour = "gray",
             linewidth = 1)+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())+
  labs(title = "Average sleep time of mamals by order",
       x = "",
       y = "Hours")
```

## Advanced ggplot 
```{r}
head(Salaries)
glimpse(Salaries)
```

```{r}
Salaries %>% 
  ggplot(aes(x = yrs.since.phd,
             y = salary,
             color = rank))+
  geom_point()+
  theme_bw()

Salaries %>% 
  ggplot(aes(x = yrs.since.phd, y = salary))+
  geom_jitter(aes(colour = rank, shape = discipline))+
  geom_smooth(method = lm)+
  facet_wrap(~sex)+
  theme_bw()+
  labs(title = "Salary vs years since PhD",
       x = "Years since PhD",
       y = "Income",
       colour = "Position",
       shape = "Research Area")

Salaries %>% 
  filter(salary < 150000) %>% 
  ggplot(aes(x = rank, y = salary, fill = sex))+
  geom_boxplot(alpha = 0.5)+
  theme_bw()+
  labs(title = "Faculty Salary by Rank and Gender",
       x = "",
       y = "",
       fill = "Gender")

Salaries %>% 
  filter(salary < 150000) %>% 
  ggplot(aes(x = rank, y = salary, fill = sex))+
  geom_boxplot(alpha = 0.5)+
  scale_x_discrete(breaks = c("AsstProf",
                              "AssocProf",
                              "Prof"),
                   labels = c("Assistant\nProfessor",
                              "Associate\nProfessor",
                              "Full\nProfessor"))+
  scale_y_continuous(breaks = c(50000, 100000, 150000, 200000),
                     labels = c("$50K", "$100K", "$150K" ,"$200K"))+
  theme_bw()+
  labs(title = "Faculty Salary by Rank and Gender",
       x = "",
       y = "",
       fill = "Gender")

Salaries %>% 
  filter(salary < 150000) %>% 
  ggplot(aes(x = rank, y = salary, fill = sex))+
  geom_boxplot(alpha = 0.5)+
  scale_x_discrete(breaks = c("AsstProf",
                              "AssocProf",
                              "Prof"),
                   labels = c("Assistant\nProfessor",
                              "Associate\nProfessor",
                              "Full\nProfessor"))+
  scale_y_continuous(breaks = c(50000, 100000, 150000, 200000),
                     labels = c("$50K", "$100K", "$150K" ,"$200K"))+
  theme_bw()+
  theme(legend.position = c(0.11, 0.78))+
  labs(title = "Faculty Salary by Rank and Gender",
       x = "",
       y = "",
       fill = "Gender")

Salaries %>% 
  mutate(discipline = factor(discipline,
                             levels = c("A", "B"),
                             labels = c("Agriculture", "Biology"))) %>% 
  ggplot(aes(yrs.since.phd, salary))+
  geom_point(aes(colour = rank), alpha = 0.5)+
  geom_smooth()+
  facet_grid(sex~discipline)+
  theme_bw()+
  theme(title = element_text(colour = "steelblue", face = "bold"))+
  labs(title = "Salary vs years since PhD",
       x = "Number of years since phD",
       y = "Current salary",
       color = "position")
```

## Ggplot Colors

```{r, warning=FALSE, message=FALSE}
library(RColorBrewer)
```

```{r}
display.brewer.all(colorblindFriendly = TRUE)

mpg %>% 
  ggplot(aes(displ, hwy, color = drv))+
  geom_jitter(size = 3)+
  scale_color_brewer(palette = "Set2")+
  theme_minimal()

mpg %>% 
  ggplot(aes(displ, hwy))+
  geom_jitter(color = "steelblue",
              size = 3)+
  theme_minimal()

mpg %>% 
  ggplot(aes(displ, hwy, color = drv))+
  geom_jitter(size = 3)+
  scale_color_manual(values = 
                       c("4" = "darkblue",
                         "f" = "darkred",
                         "r" = "darkgreen"))+
  theme_minimal()

mpg %>% 
  ggplot(aes(displ, hwy, color = drv))+
  geom_jitter(size = 3)+
  scale_color_manual(values = 
                       c("4" = "#719AC0",
                         "f" = "#75B99C",
                         "r" = "#C98D71"))+
  theme_minimal()

```


## Session Information
```{r}
sessionInfo()
```

