---
title: "Data Manipulation"
author: "Ashkan"
output: 
  html_document: 
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This document shows the process of transforming, organizing, and modifying data to extract meaningful insights or fulfill specific requirements. It involves various operations and techniques applied to data, such as filtering, sorting, aggregating, merging, splitting, and transforming.

## Loading packages
```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(gapminder)
library(MASS)
library(forcats)
```

Loading both tidyverse and MASS packages causes Error if select function is used, to solve it dplyr::select was used instead of select.


## Manipulate data using the tidyverse
```{r}
head(starwars)
```

```{r}
starwars %>% 
  dplyr::select(gender, mass, height, species) %>% 
  filter(species == "Human") %>% 
  na.omit() %>% 
  mutate(height = height/100) %>% 
  mutate(BMI = mass / height^2) %>% 
  group_by(gender) %>% 
  summarise(Average_BMI = mean(BMI))
```

## Data types in R programming
```{r}

```

## Rename variables and reorder columns
```{r}
sw <- starwars %>% 
  dplyr::select(name, height, mass, gender, everything())
head(sw)

sw_1 <- starwars %>% 
  dplyr::select(name, mass, height) %>% 
  rename(weight = mass)
head(sw_1)
```

## Using the tidyverse and dplyr packages to create a new variable
```{r}
sw_2 <- starwars %>% 
  dplyr::select(name, height, mass, gender) %>% 
  rename(weight = mass) %>% 
  na.omit() %>% 
  mutate(height = height/100) %>% 
  filter(gender == "masculine" |
           gender == "feminine")
head(sw_2)

sw_3 <- starwars %>% 
  dplyr::select(name, height, mass, gender) %>% 
  rename(weight = mass) %>% 
  na.omit() %>% 
  mutate(height = height/100) %>% 
  filter(gender %in% c("masculine", "feminine")) %>% 
  mutate(gender = recode(gender,
                         masculine = "m",
                         feminine = "f")) %>% 
  mutate(size = if_else(height > 1 & weight > 75, "big", "small"))
head(sw_3)
```

## Data filtering tips using tidyverse to filter and subset data
```{r}
head(msleep)
```

```{r}
mydata_1 <- msleep %>% 
  dplyr::select(name, sleep_total) %>% 
  filter(sleep_total > 18)
head(mydata_1)
```

```{r}
mydata_2 <- msleep %>% 
  dplyr::select(name, sleep_total) %>% 
  filter(!sleep_total > 18)
head(mydata_2)
```

```{r}
mydata_3 <- msleep %>% 
  dplyr::select(name, order, bodywt, sleep_total) %>% 
  filter(order == "Primates", bodywt > 20)
head(mydata_3)
```

```{r}
mydata_4 <- msleep %>% 
  dplyr::select(name, order, bodywt, sleep_total) %>% 
  filter(order == "Primates" | bodywt > 20)
head(mydata_4)
```

```{r}
mydata_5 <- msleep %>% 
  dplyr::select(name, sleep_total) %>% 
  filter(name == "Cow" |
           name == "Dog" |
           name == "Goat")
head(mydata_5)
```

```{r}
mydata_6 <- msleep %>% 
  dplyr::select(name, sleep_total) %>% 
  filter(name %in% c("Cow", "Dog", "Goat"))
head(mydata_6)
```

```{r}
mydata_7 <- msleep %>% 
  dplyr::select(name, sleep_total) %>% 
  filter(between(sleep_total, 16, 18))
head(mydata_7)
```

```{r}
mydata_8 <- msleep %>% 
  dplyr::select(name, sleep_total) %>% 
  filter(near(sleep_total, 16, tol = 0.5))
head(mydata_8)
```

```{r}
mydata_9 <- msleep %>% 
  dplyr::select(name, conservation, sleep_total) %>% 
  filter(is.na(conservation))
head(mydata_9)
```

```{r}
mydata_10 <- msleep %>% 
  dplyr::select(name, conservation, sleep_total) %>% 
  filter(!is.na(conservation))
head(mydata_10)
```

## Explore the data

### Data frame charachteristics

```{r}
#?starwars show useful info about the data set in R
dim(starwars)
#str(starwars) very messy that is why we use glimpse
glimpse(starwars)
#view(starwars) displlay dataframe in table format in R
head(starwars)
tail(starwars)

starwars$name
#attach(starwars) # if you attach no need to say the data frame name before 
# variable

names(starwars) #name of all variables
length(starwars) # number of variables
class(starwars$hair_color) # type of this variable
length(starwars$hair_color) # number of observations in this variable
unique(starwars$hair_color) # all of uniques value for this variable
table(starwars$hair_color) # for each value how many observations

sort(table(starwars$hair_color), decreasing = TRUE)
starwars %>% 
  dplyr::select(hair_color) %>% 
  count(hair_color) %>% 
  arrange(desc(n))


view(sort(table(starwars$hair_color), decreasing = TRUE)) # show nice in R
starwars %>% 
  dplyr::select(hair_color) %>% 
  count(hair_color) %>% 
  arrange(desc(n)) %>% 
  view()
barplot(sort(table(starwars$hair_color), decreasing = TRUE))
```

### Missing data
```{r}
#is.na(starwars$hair_color) A vector about the rows and hair color
starwars[is.na(starwars$hair_color), ] #shows only the rows that hair color is N/A 
view(starwars[is.na(starwars$hair_color), ])
```

### Numeric variable
```{r}
class(starwars$height)
length(starwars$height)
summary(starwars$height)
boxplot(starwars$height)
hist(starwars$height)
```

## Clean the data

### Factors
```{r}
class(starwars$gender)
starwars$gender_fac <- as.factor(starwars$gender)
class(starwars$gender_fac)
levels(starwars$gender_fac)

#reorder levels
starwars$gender_fac <- factor((starwars$gender_fac),
                              levels = c("masculine","feminine")) 
levels(starwars$gender_fac)
```

### Select Variables
```{r}
names(starwars)

starwars %>% 
  dplyr::select(name, height,ends_with("color")) %>% 
  names()
```

### Filter variables
```{r}
unique(starwars$hair_color)

starwars %>% 
  dplyr::select(name, height,ends_with("color")) %>% 
  filter(hair_color %in% c("blond","brown") &
           height < 180)


```

### Missing data
```{r}
mean(starwars$height, na.rm = TRUE)

starwars %>% 
  dplyr::select(name, gender, hair_color, height)

# not recommended because we do not know what we miss
starwars %>% 
  dplyr::select(name, gender, hair_color, height) %>% 
  na.omit()

#show all rows with any missing values for any variables
starwars %>% 
  dplyr::select(name, gender, hair_color, height) %>%
  filter(!complete.cases(.))

#from all rows that has at least 1 missing value
#drop rows that has height missing value
starwars %>% 
  dplyr::select(name, gender, hair_color, height) %>%
  filter(!complete.cases(.)) %>% 
  na.omit(height)

starwars %>% 
  dplyr::select(name, gender, hair_color, height) %>%
  filter(!complete.cases(.)) %>% 
  mutate(hair_color = replace_na(hair_color, "none"))


```

### Duplicates
```{r}
Names <- c("Peter", "John", "Andrew", "Peter")
Age <- c(22, 33, 44, 22)

friends <- data.frame(Names, Age)

friends[!duplicated(friends),]

friends %>% 
  distinct()
```

### Recoding Variables
```{r}
starwars %>% 
  dplyr::select(name, gender) %>% 
  mutate(gender_coded = recode(gender,
                         "masculine" = 1,
                         "feminine" = 2))
```

## Data wrangling

### Rename a variable
```{r}
msleep %>% 
  rename("conserv" = "conservation") %>% 
  glimpse()
```

### Reordering Variable
```{r}
msleep %>% 
  dplyr::select(vore, name, everything() ) %>% 
  glimpse()
```

### Change variable type
```{r}
class(msleep$vore)
msleep$vore <- as.factor(msleep$vore)

glimpse(msleep)

msleep %>% 
  mutate(vore = as.character(vore) ) %>% 
  glimpse()
```

### Select variable to work with
```{r}
names(msleep)

msleep %>% 
  dplyr::select(2:4,
         awake,
         starts_with("sleep"),
         contains("wt")) %>% 
  names()
```

### Filter and arrange data
```{r}
unique(msleep$order)

msleep %>% 
  filter((order == "Carnivora"|
            order == "Primates") &
            sleep_total > 8) %>% 
  dplyr::select(name, order, sleep_total) %>% 
  arrange(-sleep_total) %>% 
  head()

msleep %>% 
  filter(order %in% c("Carnivora", "Primates") &
            sleep_total > 8) %>% 
  dplyr::select(name, order, sleep_total) %>% 
  arrange(order) %>% 
  head(15)
```

### Conditional change (if_else)
```{r}
size_of_brain <- msleep %>% 
  dplyr::select(name, brainwt) %>% 
  drop_na(brainwt) %>% 
  mutate(brain_size = if_else(brainwt > 0.01,
                              "large",
                              "small"))
head(size_of_brain, 15)
```

### Recode data
```{r}
size_of_brain %>% 
  mutate(brain_size = recode(brain_size,
                             "large" = 1,
                             "small" = 2))

```

### Reshape data
```{r}
head(gapminder)

data <- dplyr::select(gapminder, country, year, lifeExp)
head(data)

wide_data <- data %>% 
  pivot_wider(names_from = year, values_from = lifeExp)
head(wide_data)

long_data <- wide_data %>% 
  pivot_longer(2:13,
               names_to = "year",
               values_to = "lifeExp")
head(long_data)
```

## Describe and Summarise data

```{r}
# Describe the spread, centrality and variance of numeric variable

min(msleep$awake)
max(msleep$awake)
range(msleep$awake)
IQR(msleep$awake)
mean(msleep$awake)
median(msleep$awake)
var(msleep$awake)
```

```{r}
summary(msleep)
summary(msleep$sleep_total)

msleep %>% 
  dplyr::select(sleep_total, brainwt) %>% 
  summary()
```

### Create a summary table
```{r}
msleep %>% 
  drop_na(vore) %>% 
  group_by(vore) %>% 
  summarise(Lower = min(sleep_total),
            Average = mean(sleep_total),
            Upper = max(sleep_total),
            Difference = 
              max(sleep_total) - min(sleep_total)) %>% 
  arrange(Average)
  
```

### Create Contingacy table
```{r}
glimpse(Cars93)

table(Cars93$Origin)
table(Cars93$Origin ,Cars93$AirBags)

table(Cars93$AirBags, Cars93$Origin)
addmargins(table(Cars93$AirBags, Cars93$Origin),1)
addmargins(table(Cars93$AirBags, Cars93$Origin),2)

# Sum all numbers = 100
prop.table(table(Cars93$AirBags, Cars93$Origin))
prop.table(table(Cars93$AirBags, Cars93$Origin))*100

# Sum Columns = 100
prop.table(table(Cars93$AirBags, Cars93$Origin),2)*100
round(prop.table(table(Cars93$AirBags, Cars93$Origin),2)*100,2)

# Sum rows = 100
prop.table(table(Cars93$AirBags, Cars93$Origin),1)*100

Cars93 %>% 
  group_by(Origin, AirBags) %>% 
  summarise(number = n()) %>% 
  pivot_wider(names_from = Origin,
              values_from = number)
```

## factors and categorical variables
```{r}
head(gss_cat)
glimpse(gss_cat)

unique(gss_cat$race)
gss_cat %>% 
  pull(race) %>% 
  unique()

count(gss_cat, race, sort = T)
gss_cat %>% 
  count(race)

gss_cat %>% 
  pull(race) %>% 
  levels()

gss_cat %>% 
  dplyr::select(race) %>% 
  table()

# Drop factors that is not used
gss_cat %>% 
  mutate(race = fct_drop(race)) %>% 
  pull(race) %>% 
  levels()
```

```{r}
# order factor levels by the value of
# another numeriv variable
gss_cat %>% 
  drop_na(tvhours) %>% 
  group_by(relig) %>% 
  summarise(mean_tv = mean(tvhours)) %>% 
  ggplot(aes(x = mean_tv, y = relig)) +
  geom_point(size = 4)

gss_cat %>% 
  drop_na(tvhours) %>% 
  group_by(relig) %>% 
  summarise(mean_tv = mean(tvhours)) %>% 
  mutate(relig = fct_reorder(relig, mean_tv)) %>% 
  ggplot(aes(x = mean_tv, y = relig)) +
  geom_point(size = 4)
```

```{r}
gss_cat %>% 
  drop_na(age) %>% 
  filter(rincome != "Not applicable") %>% 
  group_by(rincome) %>% 
  summarise(mean_age = mean(age)) %>% 
  ggplot(aes(x = mean_age, y = rincome))+
  geom_point(size = 4)

# reverse factor
gss_cat %>% 
  drop_na(age) %>% 
  filter(rincome != "Not applicable") %>% 
  group_by(rincome) %>% 
  summarise(mean_age = mean(age)) %>% 
  mutate(rincome = fct_rev(rincome)) %>% 
  ggplot(aes(x = mean_age, y = rincome))+
  geom_point(size = 4)
```

```{r}
gss_cat %>% 
  count(marital)

gss_cat %>% 
  ggplot(aes(marital)) + 
  geom_bar()

  
# order factor levels by frequency of
# the value of the variable
gss_cat %>% 
  mutate(marital = fct_infreq(marital)) %>% 
  count(marital)

gss_cat %>% 
  mutate(marital = fct_infreq(marital)) %>%
  mutate(marital = fct_rev(marital)) %>%
  count(marital)

gss_cat %>% 
  mutate(marital = marital %>% fct_infreq() %>% fct_rev()) %>% 
  ggplot(aes(marital)) + 
  geom_bar(fill = "steelblue", alpha = 0.5) +
  theme_bw()
```

 To recode the factors use fct_recode() and fct_collapse()
 
```{r}
gss_cat %>% 
  count(relig, sort = T)

# lumping into others
gss_cat %>% 
  mutate(relig = fct_lump(relig, 2)) %>% 
  count(relig, sort = T)
```
 
```{r}
gss_cat %>% 
  filter(!is.na(age)) %>% 
  filter(marital %in% c("Never married",
                        "Married",
                        "Widowed")) %>% 
  count(age, marital) %>% 
  group_by(age) %>% 
  mutate(prop = n / sum(n)) %>% 
  ggplot(aes(x = age, y = prop, colour = marital)) +
  geom_line(linewidth = 2, na.rm = TRUE) +
  theme_minimal()

# Reordering factors by its value corresponding 
# to the largest x value
gss_cat %>% 
  filter(!is.na(age)) %>% 
  filter(marital %in% c("Never married",
                        "Married",
                        "Widowed")) %>% 
  count(age, marital) %>% 
  group_by(age) %>% 
  mutate(prop = n / sum(n)) %>% 
  mutate(marital = fct_reorder2(marital, age, prop)) %>% 
  mutate(marital = fct_rev(marital)) %>% 
  ggplot(aes(x = age, y = prop, colour = marital)) +
  geom_line(linewidth = 2, na.rm = TRUE) +
  theme_minimal()


```

## Group by and Summarise functions
```{r}
msleep %>% 
  drop_na(sleep_rem, vore) %>% 
  group_by(vore) %>% 
  summarise("Average total sleep" = mean(sleep_total),
            "Max rem sleep" = max(sleep_rem))
```
 
## Separate and Unite
```{r}
#separate year into century and year
gapminder1 <- gapminder %>% 
  separate(col = year,
           into = c("century", "year"),
           sep = 2)

#combine century and year
gapminder1 %>% 
  unite(col = date,
        century, year,
        sep = "")

```

##  manage categorial variables
```{r}
levels(gss_cat$race)

gss_cat %>% 
   dplyr::select(race) %>% 
  table()

# drop level that is not used
gss_cat %>% 
  mutate(race = fct_drop(race)) %>% 
   dplyr::select(race) %>% 
  table()

# modify factor order
gss_cat %>% 
  mutate(race = fct_drop(race)) %>% 
  mutate(race = fct_relevel(race,
                            c("White", "Black", "Other"))) %>% 
   dplyr::select(race) %>% 
  table()

# order factor by frequency of its variable
gss_cat %>% 
  mutate(marital = fct_infreq(marital)) %>% 
  mutate(marital = fct_rev(marital)) %>% 
  ggplot(aes(marital)) +
  geom_bar(fill = "steelblue")

# order factor by frequency of anothe variable
gss_cat %>% 
  group_by(relig) %>% 
  summarise(meantv = mean(tvhours, na.rm = T)) %>% 
  mutate(relig = fct_reorder(relig, meantv)) %>% 
  ggplot(aes(meantv, relig)) +
  geom_point(size = 4, color = "steelblue")

```


## Session Information
```{r}
sessionInfo()
```

